\bf Concurrent Data Assimilation for Model-guided Learning of Cardiac Potassium Channel Activities

### 2023-07-10 writing/draft_old.tex
2nd revision

	\if0\blind
	{
		\title{\bf Concurrent Data Assimilation for Modeling and Calibrating of Disease-altered Potassium Channel Kinetics in Mouse Cardiomyocytes}
		\author{Haedong Kim $^a$, Hui Yang $^a$, Andrew R. Ednie $^b$, and Eric S. Bennett $^b$ \\
		$^a$ Department of Industrial and Manufacturing Engineering, \\
		The Pennsylvania State University, University Park, PA USA \\
        $^b$ Department of Neuroscience, Cell Biology, and Physiology, \\
        Wright State University, Dayton, OH USA}
		\date{}
		\maketitle
	} \fi

@@

\begin{abstract}
Potassium channels (K\textsubscript{v}) are responsible for repolarizing the action potential in cardiomyocytes. Different K\textsubscript{v} isoforms and their currents contribute to various phases of repolarization. In the mouse ventricle, repolarization roughly consists of two kinetic components: the rapidly inactivating current I\textsubscript{Kto} (K\textsubscript{v}4.2), and the slowly inactivating currents I\textsubscript{Kslow1} and I\textsubscript{Kslow2} (K\textsubscript{v}1.5 and K\textsubscript{v}2.1). Data assimilation is used to estimate information on K\textsubscript{v} isoforms, because only the sum of their activities is measured in the form of currents (I\textsubscript{K}). Most existing studies separate K\textsuperscript{+} currents mathematically from I\textsubscript{K} recordings and make inference of kinetics by a curve-fitting decomposition procedure. However, this traditional approach has limitations to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting decomposition only relies on the shape of K\textsuperscript{+} currents without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell one-by-one individually. There is an urgent need to develop new methods that simultaneously model the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of pathological mechanisms. We propose a novel \textit{concurrent data assimilation} framework that utilizes biophysics-based computer models of K\textsubscript{v} to delineate gating kinetics in the multiple observations concurrently to cope with cellular-level dynamics. We apply the proposed method to glyco-altered I\textsubscript{K} recordings from our previous studies. Experimental results show that the proposed method can discriminate characteristics between the control and experimental groups. This proposed method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}

\noindent%
{\it Keywords:} Data assimilation, computational electrophysiology, potassium channels, kinetics modeling, glycosylation

@@

\section{Introduction}
Voltage-gated potassium channels (K\textsubscript{v}) are critical to the electrical conduction system of the heart. K\textsubscript{v} are responsible for repolarizing the cardiac action potential (AP), a change of membrane potential over time, representing the net electrical activity in a myocyte. The AP is produced by the orchestrated activities of ion channels. Voltage-gated sodium channels (Na\textsubscript{v}) initiates fast depolarization that drives the sharp upstroke of the AP. K\textsubscript{v} play antagonistic roles that repolarize the AP with the rapid inactivation of Na\textsubscript{v}. Repolarization consists of multiple phases, to which different K\textsubscript{v} isoforms contribute with fluxes of Ca\textsuperscript{2+} and Cl\textsuperscript{-} involved through designated ion channels. This complex process can be affected by even modest changes in K\textsubscript{v} activities. Aberrant repolarization can significantly affect the AP duration and the QT interval, which lead to fatal heart diseases \citep{ravens2008role}. Glycosylation is a co/posttranslational modification that is critical for protein functions including activities of K\textsubscript{v} and other voltage-gated ion channels (VGICs) \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies in diverse directions, ranging from genome-wise to proteomic/glycomic searches, have shown clear links between altered protein glycosylation and heart diseases \citep{yung2004gene,yang2015glycoproteins,miura2016glycomics,nagai2016aberrant}. Despite its importance, the underlying mechanism is still largely unknown how changes in glycosylation contribute to heart disease onset and progression. Our \textit{in-vitro} experiments showed that glycosylation regulations modulate VGIC activities and contribute to both electrical and contractile dysfunction \citep{ednie2013sialicNav1,ednie2015sialicKv,ednie2019reduced}. In addition, we verified our experimental results via \textit{in-silico} studies and estimate how changes in the lower level (VGICs) affect the functions in the higher level (cardiomyocytes), which is one of the powers of systematic computer models \citep{du2013silico,du2015statistical,du2017silico,kim2022simulation}. \textbf{Figure~\ref{fig:framework_diagram}} illustrates this feedback loop of studying heart diseases, which will be referred to as the data assimilation framework that combines experimental data with mathematical/computational models.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/overall_diagram.pdf}
    \caption{Conceptual diagram to illustrate a data assimilation framework.}
    \label{fig:framework_diagram}
\end{figure}

However, a distinctive feature of K\textsubscript{v} raises challenges in both \textit{in-vitro} and \textit{in-silico} studies: K\textsubscript{v} have various isoforms playing different roles in AP repolarization \citep{nerbonne2005molecular}, whereas dominant isoforms exist for other major VGICs \citep{abriel2010cardiac,benitah2010type}. In the mouse ventricle, repolarization is roughly delineated as two kinetic components: the rapidly inactivating current I\textsubscript{Kto}, conducted through K\textsubscript{v}4.2, and the slowly inactivating currents such as I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, conducted through K\textsubscript{v}1.5 and K\textsubscript{v}2.1 respectively. Although it is highly desirable to study individual K\textsubscript{v} isoforms to understand pathological mechanisms in the cardiac conduction system, only the collective activities as a whole can be measured in the form of currents using whole-cell voltage-clamp K\textsubscript{v} currents (I\textsubscript{K}) because of their slightly overlapping voltage-dependence of gating \citep{brouillette2004functional}. A data assimilation method, combining experimental data with mathematical/computational models, can used to estimate information on K\textsubscript{v} isoforms. Most existing studies separate individual K\textsuperscript{+} currents mathematically from I\textsubscript{K} by a curve-fitting decomposition procedure, assuming a curve of I\textsubscript{K} consists of the summation of exponential decay functions and a constant term; each represents the shape of K\textsuperscript{+} currents \citep{brunet2004heterogeneous}. However, this traditional approach is limited in its ability to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting I\textsubscript{K} decomposition only relies on the shape of K\textsuperscript{+} current without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell with different conditions (i.e., clamp voltages) one-by-one individually. These limitations not only hinder \textit{in-silico} modeling results but, in some cases, make it impossible if modeling relies on kinetic data estimated from experimental data \citep{kim2022simulation}. There is an urgent need to develop new data assimilation methods that simultaneously delineate the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of the ion channels to provide insights into clinical applications, such as drug design.

Therefore, this paper proposes a new approach of \textit{concurrent data assimilation} for investigating the kinetics of K\textsubscript{v} isoforms from multiple I\textsubscript{K} traces measured in the same cell. A schematic flowchart of the proposed framework is shown in \textbf{Figure~\ref{fig:flow_chart}B} and compared with the current approach in \textbf{Figure~\ref{fig:flow_chart}A}. First, \textit{in-silico} models of K\textsubscript{v} and their currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. It is worth mentioning that K\textsuperscript{+} currents are generated based on the mathematically simulated gating mechanism as opposed to the curve-fitting procedure, which only mimics the shape of the currents. Second, we perform a sensitivity analysis of the kinetic parameters using fractional factorial designs to identify the parameters that have significant impacts on the current generation. Several markers are defined that best represent the characteristics of the currents in various perspectives. Third, \textit{in-vitro} data of I\textsubscript{K} recordings are coupled with the proposed models by calibrating the parameters selected by the sensitivity analysis. We apply the proposed framework to glyco-altered IK recordings from our previous studies. Our data showed that reduced N-glycosylation in mouse cardiomyocytes is sufficient to cause dilated cardiomyopathy (DCM), which is the third most common cause of heart failure and the major reason for heart transplantation \citep{weintraub2017dilated}. Last, low-dimensional embedding is utilized to visually inspect the differences of the calibrated parameters in a collective way across the cells. Experimental results show that the proposed method provides reliable estimations of K\textsubscript{v} isoform kinetics of both the wild type and disease-altered groups and successfully captures cellular variability.
\begin{figure}
    \centering
    \includegraphics{figs/flow_chart.pdf}
    \caption{Flowcharts of (A) the current approach and (B) the proposed method.}
    \label{fig:flow_chart}
\end{figure}

The main advantage of this approach is that it analyzes multiple I\textsubscript{K} recordings concurrently based on the biophysical computer models, providing strong potential for modeling the kinetics and gating mechanism of the cardiac ion channels for studying heart diseases. Our contributions are summarized as follows:
\begin{itemize}
    \item We modeled the underlying kinetics of K\textsubscript{v} isoforms from whole-cell I\textsubscript{K} recordings based on biophysical principles rather than curve-fitting procedures.
    \item We developed a concurrent data assimilation routine that harnesses multiple protocols simultaneously rather than a single protocol independently one-by-one, resulting in comprehensive cellular-level modeling.
    \item The proposed approach generates separate K\textsubscript{v} models and visualize the distributions of the kinetic parameters for each cell, which allows analyzing the cellular variability. 
    \item Our method captures differences in channel kinetics given the disease-related perturbation, such as reduced glycosylation.
\end{itemize}
The proposed data assimilation framework has strong potential to serve as a novel complementary data analytical tool to \textit{in-vitro} experiments.

\section{Research Background}
\subsection{Computational Modeling of the Heart}
Computer models of the heart have been extensively studied and used in cardiology to understand the mechanism of the organ, investigate fatal diseases, and develop drugs. Models consist of mathematical equations describing biophysical properties and physiological functions compatible with experimental observations. Mathematical and computational implementations of the heart allow studying the heart in a quantitative and predictive way \citep{whittaker2020calibration}. For example, models can be used to interpolate processes not directly observed in experiments and extrapolate to novel conditions such as disease-related perturbations \citep{rodriguez2010systems}. With the advancement of sensing technology and data analysis, computational cardiology is expanding its field. Personalized heart models pave a new way for monitoring and managing the cardiac health of individuals by characterizing individual heart signals from the large-scale IoT network and calculating dissimilarities to detect potential cardiac risk \citep{yang2020network}. Hence, cardiac computer models provide a great opportunity to mitigate heart disease, the number one cause of death in the U.S. \citep{murphy2021mortality}.

The functionality of the heart to pump blood through the body is controlled by electrical signals generated in the heart itself. Hence, the electrical activities and waves spreading through the whole heart have been studied considerably. For examle, \citet{yang2016whole} a new reaction-diffusion and pattern recognition method to model spatiotemporal dynamics of electrical waves propagating throughout the heart surface. Sensing methods, such as vectorcardiogram (VCG) and electrocardiogram (ECG), measure the electrical activities of the heart. Various analytics methods have been proposed to cope with their complex nonlinear dynamics in different heart regions over time. \citet{chen2019heterogeneous} developed a heterogeneous recurrence method that characterizes spatiotemporal patterns in disease-altered multi-channel cardiac signals. Nonlinear recurrence analysis captures and quantifies topological relationships of recurring patterns in signals, which has been proven to be an effective tool for cardiac signals that are naturally recurring over time \citep{yang2010multiscale,meyers2020cross}.

However, to perform a fine-grained analysis, it is necessary to study the cellular-level activities of the heart as opposed to the organ-level signals. Voltage-gated ion channels (VGICs) are membrane proteins in cardiomyocytes that are the functional unit of the electrical activities of cardiac muscle cells. The whole-cell patch clamp recording is a powerful experimental technique for investigating the electrical properties of excitable cells such as cardiomyocytes and neurons. It allows measuring ionic currents conducted through designated VGICs and the action potential (AP). Although the laboratory technique reveals essential properties of VGICs and cardiomyocytes, the detailed gating mechanism remains ambiguous. For example, it is hard to determine rigorously how different ion channels collectively lead to changes in the AP. Computer models, coupled with experimental observations, provide integrative insights into the data, and they have become an essential part of studying the heart, as illustrated in Figure~\ref{fig:framework_diagram} \citep{winslow2011integrative}.

Hence, modeling VGICs involves a data assimilation procedure that combines experimental data with computer models. Traditionally, computer models are calibrated to characteristic curves estimated from ionic-current recordings \citep{colman2018description}. These estimation processes include curve fitting, which assumes a functional form of the characteristic curve and finds shape parameters best fit data. However, curve fitting only considers a single current recording at each time independently, so it cannot capture cellular-level dynamics from the current recordings by changes in input stimuli, i.e., voltage-clamp protocols. It raises an additional challenge for analyzing I\textsubscript{K} recordings, because one-more curve fitting is used to separate individual current traces generated by various K\textsubscript{v} isoforms. I\textsubscript{Ksum} traces are decomposed only based on the assumption of the shapes of the component currents but not the underlying biophysical gating kinetics \citep{kim2022simulation}. Therefore, there is an urgent need to develop a novel data assimilation method for modeling and calibrating K\textsubscript{v} kinetics with considering cellular dynamics.

\subsection{Altered Glycosylation and Heart Diseases}
Protein glycosylation is one of the most abundant and diverse forms of co/posttranslational modifications that impact essential protein functions, such as modulation of receptor or ion channel activities \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies has shown the association between altered glycosylation and heart diseases, such as dilated cardiomyopathy (DCM) and hypertrophic cardiomyopathy \citep{ohtsubo2006glycosylation,ednie2019reduced2}. It is reported that up to 20\% of patients with congenital disorders of glycosylation (CDG), who commonly have modest reductions in protein glycosylation, present with cardiac deficits and idiopathic DCM \citep{marques2017cardiac}. However, the underlying pathological mechanism still remains elusive.

We have investigated how regulated glycosylation contributes to heart failure in the context of electrophysiology. Electrical signaling is orchestrated activities of a variety of ion channels and transporters. Voltage-gated ion channels (VGICs) are heavily glycosylated, with ~30\% of the channel mass consisting of N-/O-linked glycans \citep{ednie2012modulation}. Glycosylation is a multi-step process and usually ends with sialic acid added. We also reported that a saturating, electrostatic effect of negatively charged sialic attached to the terminal of N-/O-glycan branches significantly altered electrical signaling in Na\textsubscript{v} \citep{ednie2013sialicNav1,ednie2015sialicNav2} as well as K\textsubscript{v} \citep{ednie2015sialicKv}. Computational modeling has been used to further investigate the functional role of reduced sialylation in Na\textsubscript{v} and K\textsubscript{v} \citep{du2015statistical, du2017silico}. However, these studies also rely on characteristic curves and statistics estimated from curve fitting.

\section{Methods}
\subsection{Computer Models of Potassium Channel Isoforms}
Mouse models have been very useful methods for studying human cardiac electric signaling \citep{nerbonne2004studying,milani2014small,kim2022irx5}. Computer models of mouse cardiomyocytes simulate important biophysical properties of ion channels and molecular dynamics in producing the AP. In this paper, the AP is modeled as a differential equation of transmembrane currents and stimulus current I\textsubscript{stim} as below in Equation~\ref{eq:ap} where $C_{m}$ is the membrane capacitance, $t$ is time.
\begin{equation}
    \label{eq:ap}
    \begin{split}
    -C_{m}\frac{dV}{dt} &= \mathrm{I}_{\mathrm{CaL}}+\mathrm{I}_{\mathrm{p}(Ca)}+\mathrm{I}_{\mathrm{NaCa}}+\mathrm{I}_{\mathrm{Cab}}+\mathrm{I}_{\mathrm{Na}}+\mathrm{I}_{\mathrm{Nab}}+\mathrm{I}_{\mathrm{NaK}}+\mathrm{I}_{\mathrm{Cl},Ca} \\
    &+\mathrm{I}_{\mathrm{Kto}}+\mathrm{I}_{\mathrm{Kslow1}}+\mathrm{I}_{\mathrm{Kslow2}}+\mathrm{I}_{\mathrm{Kss}}+\mathrm{I}_{\mathrm{Ks}}+\mathrm{I}_{\mathrm{Kr}}+\mathrm{I}_{\mathrm{K1}}+\mathrm{I}_{\mathrm{stim}}
    \end{split}
\end{equation}
There are 15 transmembrane currents in this AP model: the L-type calcium current (I\textsubscript{CaL}), the calcium pump current (I\textsubscript{p(Ca)}), the Na\textsuperscript{+}/Ca\textsuperscript{2+} exchange current (I\textsubscript{NaCa}), the calcium background current (I\textsubscript{Cab}), the fast Na\textsuperscript{+} current (I\textsubscript{Na}), the background Na\textsuperscript{+} current (I\textsubscript{Nab}), the Na\textsuperscript{+}/K\textsuperscript{+} pump current (I\textsubscript{NaK}), the Ca\textsuperscript{2+}-activated Cl\textsuperscript{-} current (I\textsubscript{Cl,Ca}), the rapidly inactivating transient outward K\textsuperscript{+} current (I\textsubscript{Kto}), the two ultra-rapidly activating delayed rectifier K\textsuperscript{+} currents (I\textsubscript{Kslow1} and I\textsubscript{Kslow2}), the non-inactivating steady-state K\textsuperscript{+} current (I\textsubscript{Kss}), the slow delayed rectifier K\textsuperscript{+} current (I\textsubscript{Ks}), the rapid delayed rectifier K\textsuperscript{+} current (I\textsubscript{Kr}), and the time independent K\textsuperscript{+} current (I\textsubscript{K1}). Detailed implementation of mouse AP and VGIC models can be found in \cite{bondarenko2014compartmentalized,asfaw2020compartmentalized}.

>>>>

	\if0\blind
	{
		\title{\bf Concurrent Data Assimilation for Model-guided Learning of Cardiac Potassium Channel Activities}
		\author{Author information is purposely removed for double-blind review}
		\date{}
		\maketitle
	} \fi

@@

\begin{abstract}
Potassium channels (K\textsubscript{v}) are responsible for repolarizing the action potential in cardiomyocytes. There is a variety of K\textsubscript{v} isoforms and corresponding currents (e.g., I\textsubscript{Kto}, I\textsubscript{Kslow1}, I\textsubscript{Kslow2}) that contribute to different phases of repolarization. Because only the sum of their activities can be measured in the form of currents (I\textsubscript{Ksum}), there is a need to decompose into individual K\textsuperscript{+} currents and their characteristics. Most existing studies separate and make inference of K\textsubscript{v} activities via curve-fitting procedures, but there are limitations such that: 1) curve-fitting decomposition only relies on the shape of K\textsuperscript{+} current traces, which does not discern the underlying kinetics and interactions; 2) I\textsubscript{Ksum} traces can only be fitted for one clamp voltage at each time, and estimated information is analyzed in a population-averaged way later. Here, we develop a novel concurrent data assimilation method that calibrates biophysics-based subject-specific computer models to decompose and delineate kinetics of K\textsubscript{v} isoforms with multiple voltage-clamp responses simultaneously. The proposed method is evaluated and validated with whole-cell I\textsubscript{Ksum} recordings from wild-type and chronically glycosylation-deficient cardiomyocytes. Experimental results show that the proposed method effectively handles multiple responses in the voltage-clamp protocol and describes the glycosylation-conferred perturbations observed experimentally to various K\textsubscript{v} isoforms. In addition, we develop a graphical-user-interface (GUI) application that provides an enabling tool to biomedical scientists. The proposed method and pertinent software are shown to have strong potential to study K\textsubscript{v} kinetics in various heart diseases.
\end{abstract}

\noindent%
{\it Keywords:} Data assimilation, in-silico modeling, model calibration, cardiac potassium channels, congenital disorders of glycosylation

@@

\section{Introduction}
Potassium channels (K\textsubscript{v}) play critical roles in the electrical conduction system of the heart, particularly in the repolarization phase of the action potential (AP). The AP is a change of membrane potential over time, representing the net electrical activity in a cardiomyocyte, and the AP shape and duration are primarily determined by K\textsubscript{v} isoforms. The ability of the heart to pump blood through the body in an appropriate rhythm is controlled by electrical signaling. Hence, even modest changes in K\textsubscript{v} activities can significantly affect the AP duration and the QT interval, which lead to fatal heart diseases \citep{ravens2008role}. As shown in Figure~\ref{fig:ap_currents}, the different phases of AP repolarization in mouse cardiomyocytes are the result of the coordinated activity of a variety of K\textsubscript{v} isoforms (e.g., K\textsubscript{v}4.2, K\textsubscript{v}1.5, and K\textsubscript{v}2.1) and their corresponding currents (e.g., I\textsubscript{Kto}, I\textsubscript{Kslow1}, and I\textsubscript{Kslow2}). There are other major voltage-gated ion channels (VGICs), such as for Na\textsuperscript{+} (Na\textsubscript{v}) and Ca\textsuperscript{2+} (Ca\textsubscript{v}), that also contribute to the AP. The Na\textsubscript{v} are primarily responsible for the AP upstroke in the depolarization phase, while the Ca\textsubscript{v} control the cellular contraction. Aberrant activities of ion channels can significantly impact the AP and lead to fatal arrhythmias. Using whole-cell voltage-clamp recording methods, only the collective activities of K\textsubscript{v} isoforms can be measured reliably as the sum of all K\textsuperscript{+} currents (I\textsubscript{Ksum}), and these K\textsubscript{v} isoforms have overlapping biophysical properties (i.e., voltage dependence of gating and kinetics), which complicate the ability to separate one type of I\textsubscript{K} from another \citep{brouillette2004functional}. K\textsubscript{v} activities can be altered in various cardiomyopathies \citep{tristani2001molecular,giudicessi2012potassium}, thus there is an urgent need to decompose I\textsubscript{Ksum} more rigorously into individual K\textsuperscript{+} currents and estimate gating kinetics of K\textsubscript{v} isoforms to understand their pathological roles by comparing their activities in healthy versus diseased hearts and cardiomyocytes.
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/ap_currents.pdf}
    \caption{(a) Ventricular action potential in mouse cardiomyocytes and (b) diverse K\textsubscript{v} isoforms and their currents contributing to the AP shape.}
    \label{fig:ap_currents}
\end{figure}

Transgenic mouse models are the most commonly used animal models that have provided insights into cardiac research despite the differences between rodents and humans; the isoforms expressed in both small and large mammals share many similarities \citep{milani2014small}. Most existing studies using mouse models separate individual K\textsuperscript{+} currents mathematically via a curve-fitting procedure \citep{costantini2005homeodomain,ednie2019reduced2,teng2022tmem65}, assuming the I\textsubscript{Ksum} current trace represents the summation of exponential decay functions (two or three, usually) and a constant term; each represents the shape of individual K\textsuperscript{+} currents. However, the ability to scrutinize the kinetics and dynamics of individual K\textsubscript{v} isoforms is limited by this traditional method because: 1) curve-fitting decomposition only relies on the shape of K\textsuperscript{+} current traces as discreet exponential functions which does not discern the underlying kinetics and interactions of K\textsubscript{v} isoforms; 2) I\textsubscript{Ksum} traces can only be decomposed for one clamp voltage at a time, and estimated information from each fitting is analyzed in a population-averaged way later. New methods are urgently needed to simultaneously delineate the underlying kinetics of multiple I\textsubscript{Ksum} traces from the same cell concurrently for a better understanding of K\textsubscript{v} channels and their roles as well as cellular variability in diseased cardiomyocytes.

This paper presents a new approach of subject-specific concurrent data assimilation for model-guided learning of K\textsubscript{v} isoforms. First, computer models of K\textsubscript{v} and their currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. It is worth mentioning that K\textsuperscript{+} currents are generated based on the mathematically simulated gating mechanism. Second, we perform a sensitivity analysis of the kinetic parameters using fractional factorial designs to identify the parameters that have significant effects on the current generation. Several markers are defined that best represent the characteristics of the K\textsuperscript{+} currents in different perspectives. Third, a calibration routine is proposed to adjust parameters selected by the sensitivity analysis to couple \textit{in-silico} models with \textit{in-vitro} data of I\textsubscript{Ksum} recordings from the same cell concurrently. Last, low-dimensional embedding is utilized to visualize calibrated parameters in a collective way across the cells. The proposed methodology is evaluated and validated by comparing I\textsubscript{Ksum} recordings on healthy control versus chronically glycosylation-deficient cardiomyocyte. Glycosylation is a co/posttranslational modification that is critical for protein functions including activities of K\textsubscript{v} and other VGICs \citep{ohtsubo2006glycosylation,ednie2012modulation}. Our in-vitro experiments showed that chronic reduction in cardiomyocyte N-glycosylation modulates VGIC activities and contribute to both electrical and contractile dysfunction \citep{ednie2013sialicNav1,ednie2015sialicKv,ednie2019reduced}. In fact, these studies also showed that reduced cardiomyocyte complex N-glycosylation is sufficient to cause dilated cardiomyopathy (DCM), which is the third most common cause of heart failure and the major reason for heart transplantation \citep{weintraub2017dilated}.

Experimental results show that the proposed method provides reliable estimations of K\textsubscript{v} currents with underlying kinetics and successfully captures cellular variability. For example, it verifies the experimental data that showed the significant reduction in the current magnitude and elongated decaying for the glycosylation-deficient cardiomyocyte group. Additionally, computational modeling estimates channel kinetics, which was not possible in the electrophysiological experiments alone. Cellular variability is visualized by low-dimensional embedding of calibration parameters into 3D space, which shows one of the powers of systematic computer models that allow one to predict how changes in VGICs impact cellular functions. Further, we develop a graphical-user-interface (GUI) application to make the proposed method accessible to biomedical scientists for investigating K\textsubscript{v}-related channelopathy that is available as online supplemental material. The proposed method shows strong potential for modeling the kinetics and gating mechanism of K\textsubscript{v} to study heart diseases. Our contributions are summarized as follows:
\begin{itemize}
    \item We model the underlying kinetics of K\textsubscript{v} isoforms from I\textsubscript{Ksum} recordings, which enables to simulate individual K\textsuperscript{+} currents based on biophysical principles.
    \item We develop a subject-specific concurrent data assimilation routine that learns I\textsubscript{Ksum} recordings measured from the same cell at multiple membrane potentials simultaneously rather than a single I\textsubscript{Ksum} trace independently to model cellular-level dynamics.
    \item The prediction uncertainty is quantified and the distribution of kinetic parameters is visualized using low-dimensional embedding.
    \item A GUI application is provided to make the proposed method an enabling tool for biomedical scientists.
\end{itemize}
The software package of the GUI application, tutorial, and reproducible MATLAB codes for modeling and analysis results are available as supplementary material.

\section{Research Background}
\subsection{Data Assimilation and Calibration of Cardiac Models}
The functionality of the heart to pump blood through the body is controlled by electrical signals generated by the organ itself. Cardiac electrophysiology has contributed significantly to our understanding of the heart and disease-related modifications, such as congenital disorders of glycosylation (CDG). Advancements in laboratory techniques allow researchers to study molecular-level activities in cardiomyocytes via measuring ionic currents conducted through VGICs. For example, whole-cell current recordings are used to show the functions of a certain gene that encodes $\alpha$ subunits of Na\textsubscript{v} in human cardiomyocytes, leading to a small but inherent and chronic risk of acquired arrhythmia \citep{splawski2002variant}; pathophysiological roles of reduced sialylation impacting Na\textsubscript{v} and K\textsubscript{v} activities in mouse cardiomyocytes \citep{ednie2015sialicNav2,ednie2015sialicKv}; modulation of O-glycosylation causing aberrant K\textsubscript{v} activities \citep{schwetz2011sialic}. However, in-vitro experiments alone are limited to investigate detailed channel activities. There is a need to assimilate experimental data to estimate information that are not able to be observed directly. Traditionally, a curve-fitting method is used to make inferences, which assumes a functional form of information to estimate (e.g., individual currents of K\textsubscript{v} isoforms) \citep{liu2011dissection}.

In contrast, computer models consist of mathematical equations describing biophysical properties and physiological functions compatible with experimental observations. Mathematical and computational modeling allows for studying the heart in a quantitative and predictive way \citep{whittaker2020calibration}. Computer models, coupled with experimental observations, provide integrative insights into the data by calibrating model parameters \citep{winslow2011integrative}. For example, models can be used to interpolate processes not directly observed in experiments and extrapolate to novel conditions such as disease-related perturbations \citep{rodriguez2010systems}. From an engineering perspective, computer models of ion channels are dynamic system simulations that represent continuously changing gating kinetics. The dynamic nature of the models complicates the calibration process. Complex nonlinear structures of cardiac computer models also make calibration difficult. We developed statistical metamodeling and sequential design \citep{du2015statistical}, nonlinear optimization algorithms \citep{du2013silico,du2017silico}, and a heuristic optimization method \citep{kim2022simulation} in our previous in-silico studies to cope with the complexity of cardiac models. However, all these studies are based on current-shape parameters via curve fitting, or characteristic curves/statistics from additional curve fitting applied to estimated currents. In addition, only population-average models are derived and calibrated via learning the entire dataset for each healthy and diseased group. Little work has been done to leverage computer models to calibrate and delineate kinetic dynamics of K\textsubscript{v} isoforms with multiple voltage-clamp responses simultaneously.

\subsection{Congenital Disorders of Glycosylation}
Protein glycosylation is one of the most abundant and diverse forms of co/posttranslational modifications that impact essential protein functions, such as modulation of receptor or ion channel activities \citep{ohtsubo2006glycosylation,ednie2012modulation}. A growing number of studies have shown the association between altered glycosylation and heart diseases, such as dilated cardiomyopathy (DCM) and hypertrophic cardiomyopathy \citep{ednie2019reduced2,ohtsubo2006glycosylation}. It is reported that up to 20\% of patients with congenital disorders of glycosylation (CDG), who commonly show modest reductions in protein glycosylation, present with cardiac deficits, including idiopathic DCM \citep{marques2017cardiac}. However, uncovering the underlying pathological mechanisms still remains elusive. We have investigated how regulated glycosylation contributes to heart failure in the context of electrophysiology. Electrical signaling is orchestrated activities of a variety of ion channels and transporters. VGICs are heavily glycosylated, with ~30\% of the channel mass consisting of N-/O-linked glycans \citep{ednie2012modulation}. Glycosylation is a multi-step process and usually ends with sialic acid added. We reported that a saturating, electrostatic effect of negatively charged sialic attached to the terminal of N-/O-glycan branches significantly altered electrical signaling in Na\textsubscript{v} \citep{ednie2013sialicNav1,ednie2015sialicNav2} as well as K\textsubscript{v} \citep{ednie2015sialicKv}. Computational modeling has been used to further investigate the functional role of reduced sialylation in Na\textsubscript{v} and K\textsubscript{v} \citep{du2015statistical,du2017silico}.

\section{Research Methodology}
\subsection{Computer Models of Potassium Channel Isoforms}
Because of the relative ease at which their genetics can be altered, mice have been useful models for studying cardiac electric signaling \citep{milani2014small,nerbonne2004studying}. In ventricular and atrial mouse cardiomyocytes, there are three major components of I\textsubscript{Ksum}: rapidly inactivating transient outward K\textsuperscript{+} currents (I\textsubscript{Kto,f} and/or I\textsubscript{Kto,s}), which are conducted through K\textsubscript{v}4.2 and K\textsubscript{v}1.4 respectively, slowly inactivating delayed rectifier K\textsuperscript{+} currents (I\textsubscript{Kslow1} and I\textsubscript{Kslow2}), which are conducted through K\textsubscript{v}1.5/K\textsubscript{v}2.1, and the non-inactivating steady-state K\textsuperscript{+} current (I\textsubscript{Kss}) \citep{xu1999four}. I\textsubscript{Kto,s} (K\textsubscript{v}1.4) is primarily found in septal cardiomyocytes \citep{bondarenko2004computer}. In this study, we only include I\textsubscript{Kto,f} in I\textsubscript{Kto}, because our in-vitro studies focused on the ventricular apex myocytes. However, I\textsubscript{Kto} can be easily modified according to the region of ventricular cardiomyocytes. We model these four dominant K\textsuperscript{+} currents: I\textsubscript{Kto}, I\textsubscript{Kslow1}, I\textsubscript{Kslow2}, and I\textsubscript{Kss}. Figure~\ref{fig:kcurrent_example}(b) illustrates the shape of the primary K\textsuperscript{+} currents and their contribution to the whole-cell I\textsubscript{K} trace (I\textsubscript{Ksum}) given the protocol in Figure~\ref{fig:kcurrent_example}(a) that applies 0 mV voltage pulse from holding potential -70 mV from time $t^{(h)}$ to $t^{(e)}$. Figure~\ref{fig:kcurrent_example}(c) and (d) show a range of voltage steps from -30 mV to 50 mV in 10 mV increments and consequent changes in I\textsubscript{Ksum}. Because of their shape, I\textsubscript{Kss} is assumed as a constant and the other currents an exponential function in the traditional data-driven curve-fitting approach. In this approach, if all four current models are included, I\textsubscript{Ksum} is defined by \eqref{eq:expl_fitting} (i.e., tri-exponential fitting) with the shape parameters such as amplitude $\mathrm{A}_{i}$ and time constant $\tau_{i}$ for $i \in \{\mathrm{Kto}, \mathrm{Kslow1}, \mathrm{Kslow2}, \mathrm{Kss}\}$. In some cases, it is reduced to a bi-exponential function combining I\textsubscript{Kslow1} and I\textsubscript{Kslow2}. 
\begin{figure}[!ht]
    \centering
    \includegraphics{figs/kcurrent_example.pdf}
    \caption{Example of voltage-clamp protocol and K\textsuperscript{+} current traces. (a) Clamp-voltage pulse of 0 mV from the holding potential -70 mV. (b) Dominant K\textsuperscript{+} currents and their contributions to I\textsubscript{Ksum}. (c) Series of clamp-voltage pulses (-30-50 mV in 10 mV increments) from the holding potential -70 mV and (d) consequent I\textsubscript{Ksum} traces.}
    \label{fig:kcurrent_example}
\end{figure}
\begin{equation}
    \label{eq:expl_fitting}
    \mathrm{I}_{\mathrm{Ksum}} = \mathrm{A}_{\mathrm{Kto}}e^{-t/\tau_{\mathrm{Kto}}} + \mathrm{A}_{\mathrm{Kslow1}}e^{-t/\tau_{\mathrm{Kslow1}}} + \mathrm{A}_{\mathrm{Kslow2}}e^{-t/\tau_{\mathrm{Kslow2}}} +  \mathrm{A}_{\mathrm{Kss}}.   
\end{equation}
